# 2. Regression and model validation

<!-- *Describe the work you have done this week and summarize your learning.* -->

<!-- - Describe your work and results clearly.  -->
<!-- - Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods. -->
<!-- - Assume the reader has no previous knowledge of your data or the more advanced methods you are using. -->

Let's learn a bit about regression and model validation in this chapter. I am here to learn that as well, so, please, don't consider it as a source of real knowledge.

Usually, it's easier to understand a new topic on an example, therefore --- here is one:

```{r}
students2014 <- read.table(file = 'data/learning2014.txt', sep=",")
```

This will read the data that was previously prepared such that we now can investigate it. 

First, what is the data there?

```{r}
dim(students2014)
```

This means that there are 166 lines and 7 columns. Ok, what's the actual data afterall?

```{r}
str(students2014)
```

As far as I understand that is the answers to the questionnaire of students who took the "Introduction to social statistics course" (the original name in Finnish, but I still don't know Finnish, so I'll use the translation provided) in the autumn of 2014. `gender` and `age` are clearly understandable. `attitude` --- some cumulative information about the attitude to learning of statistics, `deep` --- how serious is the student about the learning process, they call it "deep approach", `stra` --- strategic approach (if a student have a plan for the week and studies, being able to make good use of the time, and so on), `surf` --- surface approach (quite the opposite, something like: "I concentrate to learning just those bits of information I have to know to pass."), `points` --- exam points.

`attitude`, `deep`, `stra`, and `surf` can have values in the interval [1, 5], where, I suppose, `1` means "strongly disagree", and '5' means "strongly agree". 
I'm not sure what was the maximum available result during the exam, but if to run

```{r}
summary(students2014)
```

one can see, that in this dataset `points` have integer values in [7, 33].

As `summary()` doesn't output a bit more interesting information about the `gender` column itself, let's try

```{r}
table(students2014$gender)
```
There were 110 female students and 56 male students that finished the course.

Trying to grasp the data as a whole, let's plot it:

```{r}
library(GGally)
library(ggplot2)
p <- ggpairs(students2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```
We know that there were more female students than male, therefore the pinkish  is for females, and blueish --- males.
First column of horizontal histograms show how many females/males are of a certain age, have a certain attitute, and so forth.
First row --- boxplots --- median, quartiles (median of lower and apper parts of the dataset), maximum, minimum and outliers. It shows that median global attitude toward statistics among females is lower than among males, males in general are less strategic then females in ther approach for studies, 
